---
title: "Dimension reduction with Wingspan data"
author: "Paulius Alaburda"
date: "2021-08-07"
description: "Using data from the board game Wingspan to present dimension reduction techniques."
categories: [r, data science]
freeze: true
---

[Wingspan](https://boardgamegeek.com/boardgame/266192/wingspan) is a wonderful game. It has an accessible theme, beautiful components and is a great example of the board game hobby. What's great is that **Wingspan** also showcases over 300 unique birds. The game is amazing for families, but it's also amazing for R since all of birds can be accessed with the `wingspan` package!

## The `birds` dataset

The `birds` dataset in the `wingspan` package contains 50 variables, some of them reflect the actual characteristics of the birds - such as the wingspan, type of nest - other properties are related to the game mechanics (type of habitat, egg capacity, type of food eaten). In cases where we have a lot of variables that we would like to compress down to 2-3 dimensions for visualisation or prediction model purposes, dimension reduction techniques are used.

This blog post is going to show how to apply different dimension reduction techniques and visualise the end result in an interactive chart of the birds similar to Joshua Kunst's [Pokemon cloud](https://jkunst.com/blog/posts/2016-03-08-pokemon-vizem-all/).

```{r setup, include=FALSE}
#| eval: false

knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(wingspan)
library(tidyverse)
library(httr)
library(rvest)
library(highcharter)
library(htmltools)

birds <- birds %>%
  mutate(nest_type = replace_na(nest_type, "No nest")) %>%
  mutate(wingspan = replace_na(wingspan, 0))
```

## Prepping data for dimensionality reduction

For all of these techniques, it's best to turn the data into a matrix. Another good thing is to drop variables that do not carry information about the birds themselves, such as the name.

```{r}
#| eval: false

bird_matrix <- birds %>%
  select(-c("common_name","scientific_name","power_text","set","food_cost_mul")) %>%
  mutate(across(everything(), ~replace_na(., "Other"))) %>%
  model.matrix(~., data = .) %>%
  as.data.frame() %>%
  as_tibble() %>%
  select(-1)
```

## Principal Component Analysis (PCA)

Principal component analysis (PCA) is a linear dimension reduction technique. Without delving into specifics, a principal component (PC) is a collection of points in a coordinate space so that the first PC captures the most variance while the second captures variance orthogonal to the first one. There are methods to check whether the first few components captured the majority of the variance but I'll leave that for another blog post. PCA is best used for continuous data. If you have transcriptomics, spectrophotometry data, PCA is going to return pretty good results. If you have a dataset with continuous and categorical data, [Multiple Factor Analysis](http://factominer.free.fr/factomethods/multiple-factor-analysis.html) may be more appropriate.

```{r}
#| eval: false

pca_rs <- prcomp(bird_matrix, scale = TRUE, center = TRUE)

birds <- pca_rs %>%
  broom::augment(birds) %>%
  select(colnames(birds),.fittedPC1,.fittedPC2)

birds %>%
  filter(set == "oceania") %>%
  ggplot(aes(x = .fittedPC1,
             y = .fittedPC2,
             color = nest_type)) +
  geom_point() +
  theme_minimal() +
  ggrepel::geom_text_repel(aes(label = wingspan)) +
  labs(title = "PCA data projection",
       subtitle = "Labels are bird wingspans, colors are nest types",
       x = "PC component 1",
       y = "PC component 2",
       color = "Nest type")
```

PCA did capture the variance in wingspan between the birds but it did not form distinct clusters. Generally, I would not expect distinct clusters since birds come in all shapes and sizes but we will see that UMAP ultimately paints a better picture. As we expected previously, PCA is not the best reduction technique for our mixed dataset.

## t-SNE

t-distributed stochastic neighbor embedding (t-SNE) was developed in 2008 to embed data points into a two or three dimensional space. t-SNE works in two ways: first, it creates a probability distribution over each pair of datapoints and assigns a probability based on how similar they are. Then, it creates a second probability distribution for the points when they're already on the manifold. Then, the divergence between the two distribution is minimised. The only problem is that t-SNE has hyperparameters so the initial conditions may not return your desired result. For further reading on hyperparameter tuning and seeing t-SNE in action, this [Distill article](https://distill.pub/2016/misread-tsne/) remains the best.

```{r, message = FALSE}
#| eval: false

library(tsne)

set.seed(1)

rs <- tsne(bird_matrix,
           max_iter = 3000)

birds$tsne_1 <- rs[,1]
birds$tsne_2 <- rs[,2]

birds %>%
  filter(set == "oceania") %>%
  ggplot(aes(x = tsne_1,
             y = tsne_2,
             color = nest_type)) +
  geom_point() +
  theme_minimal() +
  ggrepel::geom_text_repel(aes(label = wingspan)) +
  labs(title = "t-SNE data projection",
       subtitle = "Labels are bird wingspans, colors are nest types",
       x = "t-SNE coordinate 1",
       y = "t-SNE coordinate 2",
       color = "Nest type")
```

## UMAP

Uniform manifold approximation and projection (UMAP) is a non-linear reduction technique. Developed in 2018, it is similar to t-SNE. Unlike t-SNE, UMAP allows interpreting distances between clusters, UMAP is computationally faster than t-SNE and so far UMAP has shown less variance with each run. The paper on the method can be found [here](https://arxiv.org/pdf/1802.03426.pdf). Overall, like t-SNE, UMAP is best used for visualising similar or dissimilar datapoints. The method is especially useful in transcriptomics, cancer research and other fields that generate terabytes of data. But there is no one stopping us from applying UMAP to our tiny dataset!

```{r, warning = FALSE}
#| eval: false

library(umap)

set.seed(1)

rs <- umap(bird_matrix)

birds$umap_1 <- rs$layout[,1]
birds$umap_2 <- rs$layout[,2]

birds %>%
  filter(set == "oceania") %>%
  ggplot(aes(x = umap_1,
             y = umap_2,
             color = nest_type)) +
  geom_point() +
  theme_minimal() +
  ggrepel::geom_text_repel(aes(label = wingspan)) +
  labs(title = "UMAP data projection",
       subtitle = "Labels are bird wingspans, colors are nest types",
       x = "UMAP coordinate 1",
       y = "UMAP coordinate 2",
       color = "Nest type")
```

Generally, the output is similar to PCA's results - the larger birds are set to the side, with the water ones grouped more closely to other water birds and the rest grouped along owls and kestrels (smaller but just as deadly!). Colorful birds also seemed to be grouped together each time I run UMAP even though their plumage wasn't encoded in the dataset. Overall, I am happy with the result as it lets me explore the "what's between X and Y" type of questions.

## Pulling birb images from Wikimedia

The one thing the `birds` dataset does not come with is images for the birds. At first, I wanted to use the original art of **Wingspan**. Natalia Rojas, the artist for Wingspan, has only published around [half of the images](https://nataliarojasart.com/wingspan-base-game-north-america/). Instead, I am going to use real bird images sourced from the Wikimedia API. All of the birds have standard Wikipedia links based on their common names, so pulling in images is going to be easy. Or so I thought.

Some of the bird names have been modified for balancing (for example, the Australian white ibis is just the Australian ibis in **Wingspan**). Also, some of the bird names in **Wingspan** are not used for the article in Wikipedia. However, the dataset contains scientific names which have not been modified.

The image sourcing is divided into two steps. First, we use the latin name to check where would Wikipedia redirect us. We save the name of the article and use that to pull the header image from the article. This saves us the trouble of recoding some of the names, dealing with special character encoding and rechecking the data whenever a new **Wingspan** expansion is released.

```{r}
#| eval: false

get_true_name <- function(x) {

  bird_common_name <- stringr::str_to_sentence(x)
  bird_common_name <- gsub("\\s","_",bird_common_name)

  rs <- GET(glue::glue("http://en.wikipedia.org/w/api.php?action=query&format=json&titles={bird_common_name}&redirects&prop=redirects&rdlimit=max"))

  content(rs)$query$redirects[[1]]$to

}

get_bird_image <- function(x) {

  bird_common_name <- gsub("\\s","_",x)

  rs <- GET("http://en.wikipedia.org/w/api.php?",
            query = list(action = "query",
                         titles = bird_common_name,
                         prop = "pageimages",
                         format = "json",
                         pithumbsize = 300
                         ))

  link <- content(rs)$query$pages[[1]]$thumbnail$source

  return(link)

}

true_names <- map(birds$scientific_name,get_true_name)
bird_urls <- map(true_names,get_bird_image)

birds$true_common_name <- unlist(true_names)
birds$image_url <- unlist(bird_urls)
```

## Building the interactive chart

The interactive chart is built using `htmltools` and `highcharter`, with most of the code borrowed from Joshua Kunst's Pokemon post. `highcharter` is a great library that I have minimal experience with, so it's all about learning from great examples.

```{r}
#| eval: false

xt <- c("common_name", "scientific_name","wingspan","nest_type","egg_capacity")

yt <- str_c("{point.", xt, "}")

xt <- str_replace(xt, "_", " ")
xt <- str_to_sentence(xt)

tb <- tooltip_table(
  xt,
  yt,
  cellspacing="0",
  cellpadding="0",
  style = "border:none; border-collapse: collapse;"
  )

tb <- tagList(
  tags$img(
    src = "{point.image_url}",
    width = "150px", height = "125px")
  ) %>%
  as.character() %>%
  str_c(tb, .)
```

```{r}
#| eval: false

highchart() %>%
  hc_chart(zoomtype = "xy") %>%
  hc_xAxis(
    visible = FALSE,
    minRange = diff(range(birds$umap_1))/5
  ) %>%
  hc_yAxis(
    visible = FALSE,
    minRange = diff(range(birds$umap_2))/5
  ) %>%
  hc_tooltip(
    useHTML = TRUE,
    borderRadius = 0,
    borderWidth = 5,
    headerFormat = "",
    pointFormat = tb,
    footerFormat = ""
  ) %>%
  hc_add_series(
    birds,
    "point",
    hcaes(umap_1,
          umap_2,
          name = common_name,
          group = nest_type,
          color = nest_type),
    states = list(
      hover = list(
        halo = list(
          size  = 50,
          attributes = list(opacity = 1)
          )
        )
      )
    ) %>%
   hc_add_series(
    birds %>% select(umap_1, umap_2, nest_type) %>% mutate(z = 1),
    "bubble",
    minSize = 100,
    showInLegend = FALSE,
    hcaes(umap_1, umap_2, color = nest_type),
    marker = list(fillOpacity = 0.05, lineWidth = NULL),
    zIndex = -3,
    enableMouseTracking = FALSE
    )
```

The result looks excellent - the larger birds are set to the side, with the water ones grouped more closely to other water birds and the rest grouped along owls and kestrels (smaller but just as deadly!). Colorful birds also seemed to be grouped together each time I run UMAP even though their plumage wasn't encoded in the dataset. Overall, I am happy with the result as it lets me explore the "what's between X and Y" type of questions.
