---
title: "Micro-improvement for window functions"
description: "A very small change leads to a very small improvement"
author: "Paulius Alaburda"
date: 2025-01-19
categories:
  - SQL
draft: false
---

The other day I was fiddling with a silver data layer at our data warehouse to ensure that the tables are built reasonably fast. We have one table that is made up of a few joined tables and a few window functions. It was clear that building a joined table and then adding the windowed columns afterwards was faster. What surprised me was that adding the window functions was faster by building them in a CTE and joining to the original table. Apparently you can replicate this in DuckDB:

```{r}
#| cache: false

library(dplyr)
library(tidyr)
library(lubridate)
library(duckdb)
library(ggplot2)


generate_intervals <- function(date_from = "2024-07-01",
                               groups = 3,
                               duration = 10,
                               gaps = 1:2,
                               n = 10) {

  data.frame(id = 1:(n*groups),
             uuid_1 = uuid::UUIDgenerate(n*groups),
             uuid_2 = uuid::UUIDgenerate(n*groups),
             uuid_3 = uuid::UUIDgenerate(n*groups),
             uuid_4 = uuid::UUIDgenerate(n*groups),
             group_id = as.character(rep(1:groups,each = n))) %>%
    group_by(group_id) %>%
    mutate(duration = sample.int(duration, n, replace = TRUE),
           gap = c(0,sample(gaps, n-1, replace = TRUE)),
           date_kernel = as.Date(date_from)) %>%
    mutate(lag_duration = replace_na(lag(duration),0),
           cumulative_duration = cumsum(duration),
           cumulative_gap = cumsum(gap)) %>%
    mutate(cumulative_lag_duration = cumsum(lag_duration)) %>%
    mutate(date_from = date_kernel %m+% days(cumulative_gap+cumulative_lag_duration),
           date_to = date_kernel %m+% days(cumulative_gap+cumulative_duration))

}

intervals_10m <- generate_intervals(date_from = "2024-07-01", groups = 10000, duration = 10, gaps = 0, n = 100)

con <- dbConnect(duckdb::duckdb(), ":memory:")

dbWriteTable(con, "intervals", intervals_10m)

```

```{r}

interval_benchmark <- bench::mark(window_joined = dbGetQuery(con, "with tmp as (
                 select
                    id,
                    lead(date_from,1) over (partition by group_id order by id) as lead_date_from
                 from intervals)
                 select
                    intervals.*,
                    tmp.lead_date_from
                 from intervals
                 left join tmp
                  on tmp.id = intervals.id"),
            window_together = dbGetQuery(con, "select
                    intervals.*,
                    lead(date_from,1) over (partition by group_id order by id) as lead_date_from
                 from intervals"),
            check = FALSE)

autoplot(interval_benchmark)

```






