---
title: "Create a Date Dimension"
author: "Paulius Alaburda"
date: "2023-06-29"
categories: [data warehouse]
---

# What's a date dimension anyway?

I have been searching for a clean solution to building a date dimension so here I am writing my own. In R!

If you don't speak Kimballese, a date dimension is a calendar. It contains a row for each day in a year as well as attributes about those dates. For example, a date dimension for 2023 would contain 365 rows and have columns such as week number, day of week, whether the day is a holiday etc. There's really no correct way of building a date dimension as different organisations might have different ways of aggregating data across time but it is useful to have a "starter" table.

# Building the date table

First, we'll build the table using `lubridate`. My locale is Lithuanian, but your output should adapt to your locale:

```{r}

library(lubridate)
library(tibble)

date_dimension <- tibble(date = seq(as.Date('2011-01-01'),as.Date('2011-01-31'),by = 1),
                         full_date_description = format(date, format="%Y m. %B %d d."),
                         day_of_week = wday(date, label=FALSE, week_start = 1),
                         day_of_week_name = wday(date, label=TRUE, abbr = FALSE),
                         calendar_iso_week = isoweek(date),
                         calendar_week = week(date),
                         calendar_month = month(date),
                         calendar_month_name = month(date, label = TRUE, abbr = FALSE),
                         calendar_quarter = quarter(date),
                         calendar_quarter_name = paste0("Q",quarter(date)),
                         calendar_year = year(date),
                         is_weekday = as.integer(wday(date, week_start = 1) < 6))

head(date_dimension)

```

If this were a typical date dimension post, all you would have to do is adjust the date range and either export the table or write to a database. But this date dimension could be improved:

* It doesn't contain bank holidays
* If your use case is different enough from mine, you would have to edit more than the date range to get a desired result.

# Let's add holidays

To get bank holidays, you have a few options:

* [Nager.Date](https://date.nager.at/Api) - free to use, no sign up required
* [abstract](https://www.abstractapi.com/api/holidays-api) - free tier available but sign up required
* [Calendarific](https://calendarific.com/) - free tier available but sign up required
* [The Holiday API](https://holidayapi.com/) - free for previous years but paid for current and future years
* Scrape [Public Holidays Global](publicholidays.com) - free to use but I haven't checked their scraping policy

Let's use Nager.Date for this exercise. The API returns a list of dates that are bank holidays with quite a few useful attributes but I am keeping just the English and local holiday names.

```{r}

get_holidays <- function(country_code = "lt", year = "2023") {

  rs <- httr::GET(glue::glue("https://date.nager.at/api/v3/publicholidays/{year}/{country_code}"))

  out <- httr::content(rs) %>%
    tibble::enframe() %>%
    dplyr::select(value) %>%
    tidyr::unnest_wider(value) %>%
    dplyr::transmute(date = as.Date(date),
           holiday_name_local = localName,
           holiday_name = name)

  return(out)

}

get_holidays()

```

# Make it a function

To turn the code above into a function we are making the following changes:

* Pulling start and end dates, date format, week start number and country code as function arguments
* Passing the years from the date dimension to the get_holiday function. I want to create tables that span multiple years so we're using purrr::map_df to get holidays for multiple years.

```{r}

date_dimension <- function(date_from, date_to, full_date_format = "%Y m. %B %d d.", week_start = 1, country_code = "LT") {

  date_skeleton <- tibble(date = seq(as.Date(date_from),as.Date(date_to),by = 1),
                          full_date_description = format(date, format=full_date_format),
                          day_of_week = wday(date, label=FALSE, week_start = week_start),
                          day_of_week_name = wday(date, label=TRUE, abbr = FALSE),
                          calendar_iso_week = isoweek(date),
                          calendar_week = week(date),
                          calendar_month = month(date),
                          calendar_month_name = month(date, label = TRUE, abbr = FALSE),
                          calendar_quarter = quarter(date),
                          calendar_quarter_name = paste0("Q",quarter(date)),
                          calendar_year = year(date),
                          is_weekday = as.integer(wday(date, week_start = 1) < 6))

  holidays <- purrr::map_df(unique(date_skeleton$calendar_year), ~get_holidays(country_code = country_code, year = .))

  out <- date_skeleton %>%
    dplyr::left_join(holidays, by = "date") %>%
    dplyr::mutate(is_workday = is_weekday*is.na(holiday_name))

  return(out)

}

date_dimension("2024-01-01","2024-05-01")

```

That's it! If you want to learn more about date dimensions, be sure to read Kimball's Data Warehousing Toolkit Chapter 3.
